Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation

### 2025-04-29 manuscript/manuscript.qmd
update oc6 priors

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigator
      - Software
      - Visualization
      - Project administration
    affiliations:
      - Independant Consultant
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Co-Investigator
    affiliations:
      - NASA
abstract: Placeholder
plain-language-summary: Placeholder
keywords: []
key-points:
  - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: references.bib
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---

@@

### Prior Elicitation and Model Formulation

@@

###

>>>>

---
title: "Shifting paradigms in Ocean Color: Bayesian Inference for Uncertainty-Aware Chlorophyll Estimation "
format:
    agu-pdf:
        keep-tex: true
    agu-html: default
author:  
  - name: Erdem M. Karaköylü
    orcid: 0000-0002-6156-1720
    corresponding: true
    email: erdemk@protonmail.com
    roles:
      - Investigator
      - Software
      - Visualization
      - Project administration
    affiliations:
      - Independant Researcher
  - name: Susanne E. Craig
    orcid: 0000-0002-8963-0951
    corresponding: false
    email: susanne.e.craig@nasa.gov
    roles:
      - Co-Investigator
    affiliations:
      - NASA/UMBC
abstract: Placeholder
plain-language-summary: Placeholder
keywords: ['Bayesian modeling', 'Ocean Color', ]
key-points:
  - Bayesian models have not been fully adopted in Ocean Color remote sensing yet.
  - A Bayesian workflow is a principled guide to a fully Bayesian study
  - 
bibliography: references.bib
citation:
  container-title: Geophysical Research Letters
keep-tex: true
number-sections: false
date: last-modified
---

@@

We performed the following five steps. (1) Prior elicitation and model formulation; (2) Data preprocessing and exploratory analysis; (3) model fitting and diagnostics; (4) model evaluation and comparison.

### Prior Elicitation, Model Formulation, and Prior Predictive Checks

Skip to content
You said:
Here is my introduction to my Bayesian modeling study and the beginning of my methods section. Suggest a continuation to the prior elicitation section

### The Bayesian workflow
The Bayesian workflow is a recognizable and unifying approach to statistical modeling, primarily championed by statisticians at Columnbia University [@gelman_bda] subsequently endorsed and further developed by researchers dealing with scant and/or noisy data (e.g. [@mcelreath_stat_rethink]). Some researchers have modified this  researcher's field. [@wolkovich2024fourstepbayesianworkflowimproving] has proposed a four-step process for ecological modeling. 
However there are some core steps that should be followed. (1) Begin with one, or better yet,  more than one conceptual models and code them up.  Model building include prior formulation to encode existing knowledge. Priors are part of the model structure and make the researcher's assumptions transparent, providing a first avenue of critique and improvement. (2) Check assumptions by a simulation process known as *Prior Predictive Checks*. Bayesian models are generative, meaning they can run on empty, that is produce results without data. This allows a sanity check of the built-in assumptions. Non-sensical simulation results indicate assumptions must be revisited. (3) Collect data and conduct exploratory data analysis. Knowing the data will be critical to understanding potential problems during fitting; e..g multicollinearity of input features. (4) Diagnostics; the fitting process and the resulting posterior distribution provide rich  constructs that can be mined for a great deal of information. Most modern packages (e.g. Stan, PyMC, Numpyro, etc) offer a great many tools to extract insights. (5) Model compasison and selection. Often a model will perform markedly better, which is useful in terms predictive performance, but more importantly in terms of generating insights. (6) Sequential update; a model's output, the posterior distribution becomes the new prior every time new data becomes available for a new sequence of model fitting.

In this paper we leverage past work and recast $OC6$ [@oreilly2019] as a Bayesian mdoel, because out of the $OCx$ family,  it integrates the most bands (i.e. information) into its structure. In addition, we also propose two more models. (1) Bayesian Additive Regression Trees (BART, [@chipman2010bart]); a robust flexible approach successfully used for diverse applications ([@linero2017].), and insensitive to the known multicollinearity issue between ocean color bands. (2) A hierarchical partial pooling linear model inspired from [@gordon1983]'s original band ratio-based linear regression, formulated to circumvent discontinuities mentioned by [@oreilly1998], while insuring maximal use of available information contained in the data set.

We performed the following five steps. (1) Prior elicitation and model formulation; (2) Data preprocessing and exploratory analysis; (3) model fitting and diagnostics; (4) model evaluation, comparison, and selection.

### Prior Elicitation 
Before fitting the Bayesian model to observational data, we performed a prior predictive evaluation based on simulated mock inputs to ensure that our model structure and prior choices were physically reasonable and aligned with established ocean color variability. For all 3 model types we simulated chlorophyll a with approximately log-normal distribution according to [@campbell1995].
For $OC6$ and our hierarchical linear model we generated maximum band ratio (MBR) that roughly agree with [@oreilly2019] and references therein. BART prior is bases on model structured and elicitation is explained further below.

#### Mock MBR and Chlorophyll Data

We generated synthetic MBR inputs based on empirical knowledge of global ocean conditions. Specifically, the dynamic range of MBR values was assumed to span approximately 0.3 to 30, corresponding to $log_{10}(MBR)$ values between -0.5 and 1.5 [@oreilly2019]. $log_{10}(MBR)$ values were drawn from a truncated normal distribution centered at 0.75 with a standard deviation of 0.5, ensuring denser sampling in typical oligotrophic to mesotrophic regimes while allowing coverage of extreme cases. 
The corresponding predicted $log_{10}(Chl_a)$ ⁡ values were sampled, according to the physical (detection limits) and ecological expectations of chlorophyll variability [@campbell1995] from a truncated Gaussian likelihood bounded between -3 and 3.3, corresponding to chlorophyll concentrations ranging from 0.001 to 2000 mg m⁻³.

#### Model Formulation
To implement a robust Bayesian approach, we first conducted a thorough elicitation of priors informed by domain knowledge from previously published literature, where applicable. The choice of distribution for the model parameters is borne out of the following assumptions. 
1. Parameter domain is $(- ∞, ∞)$ though, per [@oreilly2019], parameters are expected to be relatively close to 0
2. Priors should be regularizing. 
3. Parameter variance is finite.

Accordingly for the $OC6$ and hierarchical models we chose weakly informative Gaussian priors. For the likelihood, we chose a truncated Gaussian defined over a domain The standard deviations for these priors were chosen conservatively large enough to allow for flexibility, but sufficiently tight to capture known realistic parameter ranges derived from extensive empirical validation.

For the BART model, priors were selected following recommendations by [@chipman2010bart] and subsequent applied literature [@linero2017; @hill2020bayesian]. The BART algorithm naturally encodes regularization and shrinkage through its tree structure, inherently balancing bias and variance. However, hyperparameters controlling tree depth and the number of trees were critical. We specified prior distributions for these hyperparameters based on guidelines from simulation studies and domain knowledge, starting with defaults recommended by the original authors and subsequently refining through preliminary runs and prior predictive checks.

We carried out formal prior predictive checks to validate these choices. By generating synthetic data from the prior distributions and inspecting the resulting predictions for chlorophyll-a, we ensured that our priors neither produced unrealistic extremes nor overly constrained the model predictions. This procedure also allowed stakeholders with domain expertise to visually assess the suitability of our priors before fitting the model to actual data, thereby promoting transparency and interpretability.

Finally, the resulting priors and model formulation and prior predictive cheks are available as a jupyter notebook in the accompanying repository for full reproducibility.

@@

### 2025-04-29 manuscript/manuscript.tex
update oc6 priors

\authors{Erdem M. Karaköylü\affil{1}, Susanne E. Craig\affil{2}}
\affiliation{1}{Independant Consultant, }\affiliation{2}{NASA, }
\correspondingauthor{Erdem M. Karaköylü}{erdemk@protonmail.com}

@@

Satellite ocean color observations have long been fundamental for
monitoring marine ecosystems, as they enable the global estimation of
chlorophyll‑a (\(Chl_a\)) ---a key indicator of phytoplankton biomass
and ocean productivity. Early empirical algorithms, notably developed by
O'Reilly et al. (John E. O'Reilly et al., 1998; John E. O'Reilly et al.,
2000), established the \(OCx\) family (where \(x\) denotes the number of
bands used) of polynomial regression models. These models relate
blue-to-green reflectance ratios (after log‑transformation) to in situ
\(Chl_a\), employing either straight band ratios (BR) or maximum band
ratios (MBR)---the latter selecting the highest available blue-to-green
ratio for any given observation as input to a high‑order polynomial.
These formulations have served as the operational foundation for
chlorophyll‑a products across a broad range of satellite ocean color
sensors---from the pioneering Coastal Zone Color Scanner (CZCS) through
SeaWiFS, MODIS, and MERIS to more recent missions---offering a
straightforward and robust approach for Case‑1 waters. However, their
performance is more limited in optically complex Case‑2 waters and
remains sensitive to atmospheric correction errors.

@@

\subsection{}\label{section}

>>>>

\authors{Erdem M. Karaköylü\affil{1}, Susanne E. Craig\affil{2}}
\affiliation{1}{Independant Researcher, }\affiliation{2}{NASA/UMBC, }
\correspondingauthor{Erdem M. Karaköylü}{erdemk@protonmail.com}

@@

Satellite ocean color observations have long been fundamental for
monitoring marine ecosystems, as they enable global estimation of
chlorophyll‑a (\(Chl_a\)) --- a key indicator of phytoplankton biomass
and ocean productivity. Early empirical algorithms, notably developed by
O'Reilly et al. (John E. O'Reilly et al., 1998; John E. O'Reilly et al.,
2000), established the \(OCx\) family (where \(x\) denotes the number of
bands used) of polynomial regression models. These models relate
blue-to-green reflectance ratios (after log‑transformation) to in situ
\(Chl_a\), employing either straight band ratios (BR) or maximum band
ratios (MBR)---the latter selecting the highest available blue-to-green
ratio for any given observation as input to a high‑order polynomial.
These formulations have served as the operational foundation for
chlorophyll‑a products across a broad range of satellite ocean color
sensors---from the pioneering Coastal Zone Color Scanner (CZCS) through
SeaWiFS, MODIS, and MERIS to more recent missions---offering a
straightforward and robust approach for Case‑1 waters. However, their
performance is more limited in optically complex Case‑2 waters and
remains sensitive to atmospheric correction errors.

@@
